---
title: "Terrestrial Ecosystem Protection Level 2024 - including combined invasive plant data"
author: "SANBI"
date: June 2025
format: 
 html:
  toc: true
  toc-location: left
  embed-resources: true
editor: visual
execute: 
  eval: false
---

### Approach

Terrestrial Ecosystem Protection Level assessment with the inclusion of additional steps to utilise combined invasive alien plant data. This assessment is for 2018 & 2024 and reveals the degree to which Protected Area targets have been met for each terrestrial ecosystem type (ecosystem representaiton). The analysis includes steps to address the effectiveness of PAs by down weighting the contribution to targets for areas which are heavily invaded by alien plants. .

```{r}

# Add packages: 
library(terra)
library(tidyverse)
library(PROJ) # seems to be required due to a conflict with postgres on some machines
 
```

### Data sources & import:

The 7 class South African National Land Cover data set for 2018 and 2022 (prepared by SANBI based on National Department of Forestry, Fisheries and the Environment SACALC data) was combined with invasive alien plant data from various sources described in the following workflow: [INV_terr /Invasives_combined.qmd](https://github.com/askowno/INV_terr/blob/main/Invasives_combined.qmd), and resulting in a 8 class national land cover.

This 8 class land cover was cross tabulated with the National Vegetation Map 2024 version, vector feature geodatabase (ESRI file geodatabase) (released January 2025) curated by SANBI [@rutherford2006; @Dayaram2019] and the Protected Areas vector feature geodatabase prepared by SANBI, built from Protected and Conserved Areas (PACA) database curated by South Africa's National Department of Forestry Fisheries and the Environment. It is a flattened spatial layer with no overlaps (PACA is not flattened and has overlapping features representing each legal declaration and declaration, and separate features for additional assignments such as World Heritage Site status). This data also includes numerous defacto PAs, the information for which was provided by government Conservation Authorities. PACA 2024Q3 (released March 2025) data used as initial source.

```{r}
#Add land cover data  (note all LC products use  WGS84 Albers Equal Area, central meridian = 25; std parallel1 = -24; std parallel2 = -33). Use Tif version as this is processed faster by R terra.
lc2022_inv <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/lc2022_inv_max2.tif")
lc2018_inv <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/lc2018_inv_max2.tif")

# add vegetation data
veg <- vect("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/NLC_Change.gdb", "NVM2024Final_IEM5_12_07012025")  #ensure it uses proj =  wgs84 aea cm 25 -24 -33

# to help differentiate masked areas (LES and SWA) from NON TERRESTRIAL areas add "NOTVEG" to T_MAPCODE where T_NAME = "Non-Terrestrial*"
veg$T_MAPCODE <- ifelse(veg$T_Name == "Non-terrestrial (estuarine habitat)" | veg$T_Name == "Non-terrestrial (Aquatic)", "NOTVEG", veg$T_MAPCODE) 

vegrast <- rasterize(veg,lc2022_inv,"T_MAPCODE")

#add PA data
pa <- vect("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/NLC_Change.gdb", "SANBI_PA_2024Q4_Mar2025") #ensure it uses proj =  wgs84 aea cm 25 -24 -33 
```

### Mask out Lesotho and Eswatini

To ensure the statistics on land cover change are applicable to South Africa the countries of Lesotho and Eswatini (which are covered by the input LC and VEG data above) need to be masked out. For RLE processes these are left in the workflow as the statistics need to be applicable to whole vegetation types.

```{r}
# import a polygon  of  Eswatini and Lesotho then make a raster mask from this and apply to vegmap so the stats generated are National Statistics

esw_les <- vect("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/NLC_change.gdb", "mask_esw_les") #ensure it uses proj =  wgs84 aea cm 25 -24 -33

mask_esw_les <- rasterize(esw_les, vegrast, "CODE")
mask_esw_les <- (mask_esw_les =="LeS") # had to do this as just using mask raster does not work for some reason

vegrast_masked <- mask(vegrast, mask_esw_les, maskvalue=TRUE) 

```

### Data preparation:

Subset PA data for 2018 and 2024 using DECL_YEAR, excluding properties degazetted using UNDECL_YEA. In preparation for this selection DECL_YEAR is set to 0 when unknown, and UNDCL_YEA is set to 9999 for those PAs that have never been degazetted.

Rasterize PA layers at each time point, using PA_SUBTYPE attribute, and snapping to the land cover grid.

```{r}

#Data preparation

# Create PA vectors for each time point
pa2018 <- subset(pa, pa$DECL_YEAR <  2019 & pa$UNDECL_YEA >2018)
pa2024 <- subset(pa, pa$DECL_YEAR <  2025 & pa$UNDECL_YEA >2024)
# Create PA rasters
pr2018 <- rasterize(pa2018,lc2018_inv,"SUBTYPE") 
pr2024 <- rasterize(pa2024,lc2022_inv,"SUBTYPE") 

```

### Analysis:

*Cross-tabulate land cover (including invasive plant data) with vegetation map and protected areas.*

Stack the land cover for each time period with the corresponding PA data and the vegetation data.

Cross-tabulate the raster stack. It is crucial to use the NA values in the cross tabulation as this will help with tracking LC outside of the PA network - this is because the PA vector and hence the raster is not continuous (not a wall to wall map) and there is no data between PA features.

```{r}

# Step 3. Cross tabulate vegetation, land cover (with invasives) and PAs. Align the land cover time point with the PA time point as far as possible (2018 land cover for 2018 PA; 2022 land cover for 2024 PA).
 
tb2018 <- as_tibble(crosstab(c(lc2018_inv,vegrast_masked,pr2018),useNA=TRUE))
tb2018 <- tb2018 %>% add_column(YEAR =2018) %>% rename(LC = lc2018_inv_max2) 
write.csv(tb2018, file = "outputs/pa2018_inv2.csv")

tb2024 <- as_tibble(crosstab(c(lc2022_inv,vegrast_masked,pr2024),useNA=TRUE))
tb2024 <- tb2024 %>% add_column(YEAR =2024) %>% rename(LC = nlc2022_7class) 
write.csv(tb2024, file = "outputs/pa2024_inv2.csv")

```

*Prepare analysis tables*

Calculate the required per ecosystem type metrics: total extent, extent within PA, extent that is in natural state within PA, and the extent of the natural area in the PA that is heavily invaded with invasive alien plants. The heavily invaded areas are typically \>50% canopy cover of IAP and are considered severely modified (\>70% biotic disruption / modification). As such, the areas within the PA system that are invaded do not contribute fully to achieving the representivity targets (national and global). Their contribution is down-weighted by 70%.

*Natural extent within PA + (0.7 x invaded extent within PA) = Contribution to targets using in EPL calculations.*

**EPL classes:**

-   Well Protected (WP); target met or exceeded

-   Moderately Protected (MP); 50-99% of target met

-   Poorly Protected (PP); 5-49% of target met

-   Not Protected (NP); less than 5% of target met

*National targets are ecosystem specific and range from 15-31%*

*GBF target is 30%*

#### Calculate EPL 2024

```{r}
# Build up the tables  required to compute EPL2024 

options(scipen = 999)
# clean up the results of the cross tabulation 
tb2024 <- read.csv("C:/Rdata/EPL_terr/outputs/pa2024_inv2.csv") # read in data to avoid cross tab
tb2024 <- tb2024 %>%
  filter(n != 0, !is.na(LC), LC != "NaN", LC != 0,!is.na(T_MAPCODE)) %>% 
  mutate(SUBTYPE = if_else(SUBTYPE == "" | is.na(SUBTYPE), "NOTPA", SUBTYPE)) %>% # Non PA pixels we want to keep for the stats
         mutate(km2 = (n*400)/1000000, 
         LC_text = ifelse(LC==1, "Nat", ifelse(LC==8, "Inv", "Notnat")),
         # extra code to add in Biome and PA Types - not required
         BIOME = str_replace_all(T_MAPCODE, 
                                  c("AT.*" = "Albany Thicket", 
                                    "AZ.*" = "Azonal", 
                                    "CB.*" = "IOCB", 
                                    "D.*"  = "Desert", 
                                    "FF.*" = "Fynbos", 
                                    "FR.*" = "Fynbos",
                                    "FS.*" = "Fynbos",
                                    "FO.*" = "Forest", 
                                    "G.*"  = "Grassland",  
                                    "NK.*" = "Nama-Karoo", 
                                    "SK.*" = "Succulent Karoo",
                                    "SV.*" = "Savanna",
                                    "NOTVEG.*" = "NOTVEG")),
         TYPE = str_replace_all(SUBTYPE, 
                                c("FNR.*" = "FNR", 
                                  "FWA.*" = "FWA", 
                                  "MCA.*" = "MCA", 
                                  "NP.*" = "NP", 
                                  "MPA.*" = "MPA",
                                  "NR.*" = "NR", 
                                  "PE.*" = "PE", 
                                  "SNR.*" = "SNR", 
                                  "WHS.*" = "WHS",
                                  "NOTPA.*" = "NOTPA")))%>%
  relocate(LC, .after = T_MAPCODE)%>%
  relocate(LC_text, .after = LC)%>%
  relocate(BIOME, .after = T_MAPCODE) %>%
  relocate(km2, .after = n)%>%
  relocate(TYPE, .after = SUBTYPE)%>%
  relocate(YEAR, .after = TYPE)

# make a datatable from the vegsa, then extract CONSTARGETS for EPL computation
tb_tar <- as_tibble(veg)
tb_tar <- tb_tar %>% 
  select(c(T_MAPCODE, T_CNSRV_TRGT, T_BIOME)) %>%
  distinct(T_MAPCODE, T_CNSRV_TRGT, T_BIOME) %>%
  filter(T_MAPCODE != "NOTVEG", T_MAPCODE != "", T_MAPCODE != "<Null>", T_MAPCODE != "Gd9") %>% # Gd9 is a veg type only occuring in Lesotho and needs to excluded from SA stats 
  rename(tar_sa = T_CNSRV_TRGT) %>%
  relocate(tar_sa, .after = T_BIOME) %>%
  mutate(tar_gbf = 30)
  
#  table of full (potential / historical) extent per ecosystem type
tb_veg <- tb2024 %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_veg = sum(km2),.groups = 'drop')

# table of ecosystem extent in PA - per type  
tb_vegpa24 <- tb2024 %>%
 filter(TYPE != "NOTPA",T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_vegpa24 = sum(km2),.groups = 'drop') 

# table of nat ecosystem (remnants) extent in PA - per type (invaded areas counted as natural at this point)
tb_vegrempa24 <- tb2024 %>%
 filter(TYPE != "NOTPA", T_MAPCODE != "NOTVEG", LC_text != "Notnat") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_vegrempa24 = sum(km2),.groups = 'drop') 

# table of invaded extent within renmants in PA - per type  
tb_invpa24 <- tb2024 %>%
 filter(TYPE != "NOTPA",T_MAPCODE != "NOTVEG", LC_text == "Inv") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_invpa24 = sum(km2),.groups = 'drop') 


# join  the tables and calculate EPL
epl2024 <- tb_tar %>%
  left_join(tb_veg, by = "T_MAPCODE") %>%
  left_join(tb_vegpa24, by = "T_MAPCODE") %>%
  left_join(tb_vegrempa24, by = "T_MAPCODE") %>%
  left_join(tb_invpa24, by = "T_MAPCODE") %>%
    mutate(across(everything(), ~ replace_na(.x, 0))) %>%
    mutate(prp_pa_base24 = ext_vegpa24/ext_veg) %>%
    # make targets km2
    mutate(ext_tar_sa = tar_sa/100 * ext_veg, 
           ext_tar_gbf = tar_gbf/100 * ext_veg) %>%
      #BASE EPL using full extent - counting all of eco in PA
  mutate(epl_base24 = ifelse(ext_vegpa24 >= ext_tar_sa, "WP", 
                             ifelse(ext_vegpa24 >= ext_tar_sa*0.5, "MP", 
                                    ifelse(ext_vegpa24 >= ext_tar_sa*0.05, "PP", "NP" )))) %>%
 #NATURAL EPL using natural extent (ignoring invasives) as per NBA 2018
   mutate(epl_nat24 = ifelse(ext_vegrempa24 >= ext_tar_sa, "WP", 
                             ifelse(ext_vegrempa24 >= ext_tar_sa*0.5, "MP", 
                                    ifelse(ext_vegrempa24 >= ext_tar_sa*0.05, "PP", "NP" )))) %>%
  #ADJUSTED EPL using natural and downweighting contribution of invaded areas 
  #make new col for adjustments
  #invaded areas downweighted by 70% (ie only 30% of extent will contribute to target achievement)
  mutate(ext_vegrempa_inv24 = ext_vegrempa24 - (ext_invpa24*0.7)) %>%
  mutate(epl_nat_inv24 = ifelse(ext_vegrempa_inv24 >= ext_tar_sa, "WP", 
                             ifelse(ext_vegrempa_inv24 >= ext_tar_sa*0.5, "MP", 
                                    ifelse(ext_vegrempa_inv24 >= ext_tar_sa*0.05, "PP", "NP" ))))
 
#export the results of the base, natural and adjusted EPL for 2024
write.csv(epl2024, file = "outputs/results_df_EPL_2024_invasives2.csv") #export results 
```

#### Calculate EPL 2018

```{r}
# Build up the tables  required to compute EPL2018 


tb2018 <- read.csv("C:/Rdata/EPL_terr/outputs/pa2018_inv2.csv") # read in data to avoid cross tab
tb2018 <- tb2018 %>%
  filter(n != 0, !is.na(LC), LC != "NaN", LC != 0,!is.na(T_MAPCODE)) %>% 
  mutate(SUBTYPE = if_else(SUBTYPE == "" | is.na(SUBTYPE), "NOTPA", SUBTYPE)) %>% # Non PA pixels we want to keep for the stats
         mutate(km2 = (n*400)/1000000,
         LC_text = ifelse(LC==1, "Nat", ifelse(LC==8, "Inv", "Notnat")),
         BIOME = str_replace_all(T_MAPCODE, 
                                  c("AT.*" = "Albany Thicket", 
                                    "AZ.*" = "Azonal", 
                                    "CB.*" = "IOCB", 
                                    "D.*"  = "Desert", 
                                    "FF.*" = "Fynbos", 
                                    "FR.*" = "Fynbos",
                                    "FS.*" = "Fynbos",
                                    "FO.*" = "Forest", 
                                    "G.*"  = "Grassland",  
                                    "NK.*" = "Nama-Karoo", 
                                    "SK.*" = "Succulent Karoo",
                                    "SV.*" = "Savanna",
                                    "NOTVEG.*" = "NOTVEG")),
         TYPE = str_replace_all(SUBTYPE, 
                                c("FNR.*" = "FNR", 
                                  "FWA.*" = "FWA", 
                                  "MCA.*" = "MCA", 
                                  "NP.*" = "NP", 
                                  "MPA.*" = "MPA",
                                  "NR.*" = "NR", 
                                  "PE.*" = "PE", 
                                  "SNR.*" = "SNR", 
                                  "WHS.*" = "WHS",
                                  "NOTPA.*" = "NOTPA")))%>%
  relocate(LC, .after = T_MAPCODE)%>%
  relocate(LC_text, .after = LC)%>%
  relocate(BIOME, .after = T_MAPCODE) %>%
  relocate(km2, .after = n)%>%
  relocate(TYPE, .after = SUBTYPE)%>%
  relocate(YEAR, .after = TYPE)

# table of eco extent in pa2018 - per type  
tb_vegpa18 <- tb2018 %>%
 filter(TYPE != "NOTPA",T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_vegpa18 = sum(km2),.groups = 'drop') 

# table of nat remnant extent in pa - per type ( inv counted as nat)  
tb_vegrempa18 <- tb2018 %>%
 filter(TYPE != "NOTPA", T_MAPCODE != "NOTVEG", LC_text != "Notnat") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_vegrempa18 = sum(km2),.groups = 'drop') 

# table of inv extent within remnants in pa - per type  
tb_invpa18 <- tb2018 %>%
 filter(TYPE != "NOTPA",T_MAPCODE != "NOTVEG", LC_text == "Inv") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_invpa18 = sum(km2),.groups = 'drop') 


# join up the tables and calculate EPL
epl2018 <- tb_tar %>%
  left_join(tb_veg, by = "T_MAPCODE") %>%
  left_join(tb_vegpa18, by = "T_MAPCODE") %>%
  left_join(tb_vegrempa18, by = "T_MAPCODE") %>%
  left_join(tb_invpa18, by = "T_MAPCODE") %>%
    mutate(across(everything(), ~ replace_na(.x, 0))) %>%
    mutate(prp_pa_base18 = ext_vegpa18/ext_veg) %>%
    # make targets km2
    mutate(ext_tar_sa = tar_sa/100 * ext_veg, 
           ext_tar_gbf = tar_gbf/100 * ext_veg) %>%
  
  #calc EPL  rounding proportions to 3 decimal places
  #BASE EPL using full extent - counting all of eco in PA
  mutate(epl_base18 = ifelse(ext_vegpa18 >= ext_tar_sa, "WP", 
                             ifelse(ext_vegpa18 >= ext_tar_sa*0.5, "MP", 
                                    ifelse(ext_vegpa18 >= ext_tar_sa*0.05, "PP", "NP" )))) %>%
 #NATURAL EPL using natural extent (ignoring invasives) as per NBA 2018
   mutate(epl_nat18 = ifelse(ext_vegrempa18 >= ext_tar_sa, "WP", 
                             ifelse(ext_vegrempa18 >= ext_tar_sa*0.5, "MP", 
                                    ifelse(ext_vegrempa18 >= ext_tar_sa*0.05, "PP", "NP" )))) %>%
  #ADJUSTED EPL using natural and downweighting contribution of invaded areas 
  #make new col for adjustments
  #invaded areas downweighted by 70% (ie only 30% of ext will contribute to target achievement)
  mutate(ext_vegrempa_inv18 = ext_vegrempa18 - (ext_invpa18*0.7)) %>%
  mutate(epl_nat_inv18 = ifelse(ext_vegrempa_inv18 >= ext_tar_sa, "WP", 
                             ifelse(ext_vegrempa_inv18 >= ext_tar_sa*0.5, "MP", 
                                    ifelse(ext_vegrempa_inv18 >= ext_tar_sa*0.05, "PP", "NP" ))))
 
#export the results of the adjusted EPL2018
write.csv(epl2018, file = "outputs/results_df_EPL_2018_invasives2.csv") #export results 
```

#### **Summarise Ecosystem Protection Levels for 2018 and 2024 per Biome & calculate EPL Index**

EPL Index calculation (Harris et al in prep)

(sum of WP \*3 + sum of MP \*2 + sum of PP \*1 + sum NP \*0 ) / sum all type \*3

*Calculate biome level summary of EPL 2024 and calculate EPLI 2024*

```{r}

#summarise EPL 2024 per biome with adjustments for invasives 
epl2024inv_sum <- epl2024 %>%
  group_by(T_BIOME,epl_nat_inv24) %>%
  summarise(count = n(),.groups = 'drop') %>%
  pivot_wider(names_from = epl_nat_inv24, 
              values_from = count, values_fill = 0) %>%
  relocate(WP, .before = MP)%>%
  relocate(NP, .after = PP) %>%
  # Add row-wise TOTAL
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  ungroup() %>%
  # Add TOTAL row (column-wise sums)
  bind_rows(summarise(., T_BIOME = "Total", across(where(is.numeric), sum))) %>%
  # Add EPLindex
  mutate(EPLI = ((WP*3)+(MP*2)+(PP*1)+(NP*0))/((WP+MP+PP+NP)*3)) 

write.csv(epl2024inv_sum, file = "outputs/results_df_EPL_2024_biome_invasives2.csv") #export results 

#summarise EPL 2024 per biome without using any adjustments for invasives (similar to NBA 2018)
epl2024nat_sum <- epl2024 %>%
    group_by(T_BIOME,epl_nat24) %>%
  summarise(count = n(),.groups = 'drop') %>%
  pivot_wider(names_from = epl_nat24, 
              values_from = count, values_fill = 0) %>%
  relocate(WP, .before = MP)%>%
  relocate(NP, .after = PP) %>%
   # Add row-wise TOTAL
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  ungroup() %>%
  # Add TOTAL row (column-wise sums)
  bind_rows(summarise(., T_BIOME = "Total", across(where(is.numeric), sum))) %>%
  # Add EPLindex
  mutate(EPLI = ((WP*3)+(MP*2)+(PP*1)+(NP*0))/((WP+MP+PP+NP)*3)) 

write.csv(epl2024nat_sum, file = "outputs/results_df_EPL_2024_biome2_natural.csv") #export results 

```

*Calculate biome level summary of EPL 2018 and EPLI 2018*

```{r}

#summarise EPL2018 per biome with adjustments for invasives 
epl2018inv_sum <- epl2018 %>%
    group_by(T_BIOME,epl_nat_inv18) %>%
  summarise(count = n(),.groups = 'drop') %>%
  pivot_wider(names_from = epl_nat_inv18, 
              values_from = count, values_fill = 0) %>%
  relocate(WP, .before = MP)%>%
  relocate(NP, .after = PP) %>%
   # Add row-wise TOTAL
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  ungroup() %>%
  # Add TOTAL row (column-wise sums)
  bind_rows(summarise(., T_BIOME = "Total", across(where(is.numeric), sum))) %>%
  # Add EPLindex
  mutate(EPLI = ((WP*3)+(MP*2)+(PP*1)+(NP*0))/((WP+MP+PP+NP)*3)) 

write.csv(epl2018inv_sum, file = "outputs/results_df_EPL_2018_biome_invasives2.csv") #export results 

#summarise EPL per biome without using any adjustments for invasives (similar to NBA 2018)
epl2018nat_sum <- epl2018 %>%
    group_by(T_BIOME,epl_nat18) %>%
  summarise(count = n(),.groups = 'drop') %>%
  pivot_wider(names_from = epl_nat18, 
              values_from = count, values_fill = 0) %>%
  relocate(WP, .before = MP)%>%
  relocate(NP, .after = PP) %>%
   # Add row-wise TOTAL
  rowwise() %>%
  mutate(Total = sum(c_across(where(is.numeric)))) %>%
  ungroup() %>%
  # Add TOTAL row (column-wise sums)
  bind_rows(summarise(., T_BIOME = "Total", across(where(is.numeric), sum))) %>%
  # Add EPLindex
  mutate(EPLI = ((WP*3)+(MP*2)+(PP*1)+(NP*0))/((WP+MP+PP+NP)*3)) 

write.csv(epl2018nat_sum, file = "outputs/results_df_EPL_2018_biome2_natural.csv") #export results 
```

#### **Make graphs for NBA**

```{r}
# to avoid rerunning spatial analysis load data from file...
#data <- read_csv("C:/Rdata/EPL_terr/outputs/results_df_EPL_2024_biome_invasives2.csv")
#data <- read_csv("C:/Rdata/EPL_terr/outputs/results_df_EPL_2018_biome_invasives2.csv")

# load nbaR package : first run this in console devtools::install_github("SANBI-NBA/nbaR") 
library(nbaR)

# Use NBA package to make EPL graph (need high res output for booklet)
EPL <- epl2024inv_sum %>%
    rename(`Well Protected` = WP, 
           `Moderately Protected` = MP, 
           `Poorly Protected` = PP, 
           `Not Protected` = NP ) # %>%
  # select(-...1) # required if importing from CSV
 
epl24_bar_plot <- nba_plot(EPL,
                  `T_BIOME`,
                  2:5,
                 CHRT = "bar",
                 NUM = TRUE,
                 LAB = "Percentage of ecosystem types",
                SAVE = NULL)

epl24_bar_plot

#export the graph
ggsave(
  filename = "outputs/epl24_bar_plot.jpeg", # File name
  plot = epl24_bar_plot,                   # Plot object
  device = "jpeg",                        # File format
  width = 16, height = 12, units = "cm",   # Dimensions
  dpi = 300                               # Resolution
)

#make a line plot of EPL Index using 2018 and 2014 timepoints - request Natasha to impve this or add to package.

EPLI <- epl2024inv_sum %>%
  select(T_BIOME, EPLI) %>% 
  rename(EPLI24 = EPLI) %>%
  left_join(epl2018inv_sum, by ="T_BIOME") %>%
   rename(EPLI18 = EPLI) %>%
  select(T_BIOME,EPLI18, EPLI24) %>%
  pivot_longer(cols = c(2,3), names_to = "Year", values_to = "EPLI") %>%
    mutate(Year = case_when(
    Year == "EPLI18" ~ "2018",
    Year == "EPLI24" ~ "2024",
    TRUE ~ Year
  ))
  
  epli_line_plot <- ggplot(EPLI, aes(x = Year, y = EPLI, group = T_BIOME, color = T_BIOME)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(x = "Year", y = "EPLI") +
  scale_x_discrete(limits = c("2018", "2024"))
  
  epli_line_plot
  
  #export the graph
ggsave(
  filename = "outputs/epli_line_plot.jpeg", # File name
  plot = epli_line_plot,                   # Plot object
  device = "jpeg",                        # File format
  width = 16, height = 12, units = "cm",   # Dimensions
  dpi = 300                               # Resolution
)
```
